---
title: "Data Science Capstone - Milestone Report"
author: "Ryan Wissman"
date: "September 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

The purpose of this document is to describe and summarize the data found within the SwiftKey text data set. Here, I will outline some brief exploratory data analysis of the blogs, twitter, and news datasets as well as provide some initial analysis. This is the first part of the Coursera Data Science capstone and is the first step in creating my prediction algorithm.   

## Loading the Dataset

First we will load the data, which was already downloaded to our computer, as well as any prerequisite libraries. The three basic files that we will use for this analysis are as follows: 

* en_US.blogs.txt
* en_US.news.txt
* en_US.twitter.txt

```{r}
#Load the libraries
library(stringi)

#Load the files
file_blogs = "F:/My Documents/GitHub/Coursera-DataScienceCapstone/SwiftKey_DataFiles/en_US.blogs.txt"
file_news = "F:/My Documents/GitHub/Coursera-DataScienceCapstone/SwiftKey_DataFiles/en_US.news.txt"
file_twitter = "F:/My Documents/GitHub/Coursera-DataScienceCapstone/SwiftKey_DataFiles/en_US.twitter.txt"

dataset_blogs <- readLines(file_blogs, encoding='UTF-8', warn=FALSE, skipNul=TRUE)
dataset_news <- readLines(file_news, encoding='UTF-8', warn=FALSE, skipNul=TRUE)
dataset_twitter <- readLines(file_twitter, encoding='UTF-8', warn=FALSE, skipNul=TRUE)
```

## Statistics - Exploring the Data

First, let's take a look at some of the basics. We will determine both the size of the files as well as the line and character counts of each.

### File Sizes

```{r}
#Get file sizes in MB
size_blogs <- file.info(file_blogs)$size / 1024 / 1024
size_news <- file.info(file_news)$size / 1024 / 1024
size_twitter <- file.info(file_twitter)$size / 1024 / 1024

#Print sizes in table for easy reading
data <- as.table(matrix(c(size_blogs, size_news, size_twitter)))
colnames(data) <- c("File Size in MB")
rownames(data) <- c("Blogs File", "News File", "Twitter File")
```

### Line Counts


## References

This report uses the SwiftKey dataset which can be downloaded here: http://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip 